---
title: "GR5065 Homework 2"
date: "Due February 9, 2021 at 8PM New York Time"
author: "Your Name"
output: 
  pdf_document: 
    latex_engine: xelatex
    number_sections: yes
urlcolor: blue
---

```{r, setup}
# call the set.seed function once here to make the knitting conditionally deterministic
```

# Equilibrium Climate Sensitivity

## Drawing from the prior

## Truncating the prior

## Describing the truncated prior

## PDF of a Ratio of Normals

```{r}
dratio <- function(z, mu_X = -3.7, mu_Y = -1.6, 
                   sigma_X = 0.2 * 3.7, sigma_Y = 0.5 * 1.6) {
  # implement
}
```

## Describing the truncated prior, part II


## The likelihood function components

```{r}
e <- c(Low_ii = .25, Low_iii = .35, Low_iv = .2, High_i = .75, High_ii = .65, High_iii = 0.6)
c <- c(Low_ii = 1.5, Low_iii = 1.5, Low_iv =  2, High_i = 4.0, High_ii = 4.5, High_iii = 4.5)
```
```{r}
likelihood <- function(chi, e_j, c_j) {
  # implement
}
```

## Posterior PDF

```{r}
numerator <- function(chi, e = e, c = c) {
  # implement
}
```

# Darts

```{r}
darts::drawBoard(new = TRUE)
```

## Drawing from a bivariate normal distribution

## Normal Prior Distributions

# Scoring Function

```{r}
score <- function(x, y) {
  stopifnot(is.numeric(x), length(x) == 1, is.numeric(y), length(y) == 1)
  
  # convert x and y in Cartesian coordinates to a radius and angle in polar coordinates
  # https://en.wikipedia.org/wiki/Polar_coordinate_system
  radius <- sqrt(x ^ 2 + y ^ 2)
  angle  <- atan2(y, x)
  
  if (radius > 170)   return(0)  # misses dartboard
  if (radius <= 6.35) return(50) # double bullseye
  if (radius <= 15.9) return(25) # single bullseye
  
  margin <- pi / 20
  interval <- margin * 2
  small <- pi / 2 - margin - 0:19 * interval
  large <- pi / 2 + margin - 0:19 * interval
  bed <- which(angle > small & angle <= large)
  if (length(bed) == 0) {
    angle <- angle - 2 * pi
    bed <- which(angle > small & angle <= large)
  }
  S <- darts:::getConstants()$S  # 20, 1, ..., 5
  score <- S[bed]
  if (dplyr::between(radius, 99, 107))       score <- 3 * score # in triple ring
  else if (dplyr::between(radius, 162, 170)) score <- 2 * score # in double ring
  return(score)
}
```

## Estimating the parameters

## Expected scores


