---
title: "GR5065 Homework 1 Answer Key"
date: "Due January 26, 2021 at 8PM New York Time"
author: "Ben Goodrich"
editor_options: 
  chunk_output_type: console
urlcolor: blue
output: 
  pdf_document: 
    latex_engine: xelatex
    number_sections: yes
header-includes:
   - \usepackage{amsmath}
   - \usepackage{color}
   - \usepackage{pst-poker}
---

```{r, setup}
set.seed(20210126) # this makes the PRNG conditionally deterministic
# but your numbers below will differ if you used a different seed
```

# Bowling

```{r}
# computes the x-th Fibonacci number without recursion and with vectorization
F <- function(x) {
  stopifnot(is.numeric(x), all(x == as.integer(x), na.rm = TRUE))
  sqrt_5 <- sqrt(5) # defined once, used twice
  golden_ratio <- (1 + sqrt_5) / 2
  return(round(golden_ratio ^ (x + 1) / sqrt_5))
}
```

## R Implementation

```{r}
Pr <- function(x, n = 10) ifelse(x > n, 0, F(x) ^ 2) / (F(n) * F(n + 1))
```

## Admissibility

Since $\mathcal{F}\left(x\right) > 0$, the ratio that defines the probability
function is always strictly positive. And since $x \leq n$, the ratio is always less than
one. It is less obvious that the probabilities add up to one over the entire sample
space, $\Omega$, but it is [true](https://en.wikipedia.org/wiki/Fibonacci_number#Combinatorial_identities),
which can be seen in this case from
```{r}
Omega <- 0:10
sum(Pr(Omega))
```

## Simulating a Game of Bowing

### First Roll

```{r}
x_1 <- sample(Omega, size = 1, prob = Pr(Omega))
```

### Second Roll

If $x_1 = 10$, then the second roll would not actually happen, but if it were to occur,
then no more pins would be knocked down because there are no pins available on the
second roll.

```{r}
x_2 <- sample(Omega, size = 1, prob = Pr(Omega, n = 10 - x_1))
```

### Probability of a Frame

The probability of these (one or) two rolls happening in a frame is
given by the General Multiplication Rule:
```{r}
Pr(x_1) * Pr(x_2, n = 10 - x_1)
```

### Game

A sequence of pins being knocked down over an entire game of bowling is:
```{r}
pins <- matrix(NA_integer_, nrow = 12, ncol = 2) # tenth frame may have 2 additional rolls
for (frame in 1:10) {
  x_1 <- sample(Omega, size = 1, prob = Pr(Omega))
  if (x_1 < 10) {
    x_2 <- sample(Omega, size = 1, prob = Pr(Omega, n = 10 - x_1))
    pins[frame, ] <- c(x_1, x_2)
  } else pins[frame, 1] <- x_1
}
if (x_1 == 10) { # strike on 10th frame
  x_1 <- sample(Omega, size = 1, prob = Pr(Omega))
  if (x_1 < 10) {
    x_2 <- sample(Omega, size = 1, prob = Pr(Omega, n = 10 - x_1))
    pins[11, ] <- c(x_1, x_2)
  } else {
    x_2 <- sample(Omega, size = 1, prob = Pr(Omega))
    pins[11:12, 1] <- c(x_1, x_2)
  }
} else if ( (x_1 + x_2) == 10 ) { # spare on 10th frame
  pins[11, 1] <- sample(Omega, size = 1, prob = Pr(Omega))
}
```

### Probability of a Game

Assuming frames are independent of each other, the probability
of this exact sequence of pins being knocked down is:
```{r}
prod(Pr(pins[ , 1], n = 10), 
     Pr(pins[ , 2], n = 10 - pins[ , 1]), na.rm = TRUE)
```

This probability is quite small, as should be anticipated when
multiplying many numbers between $0$ and $1$ together. In fact,
the single most likely sequence of frames is all strikes (i.e. a perfect game)
which has a probability of only `r Pr(10)^12` under this model. Nevertheless,
all possible sequences of pins have probabilities that sum to $1$
because we have used the General Multiplication Rule appropriately.

# Poker

Rick Schoenberg has a 
[textbook](https://www.google.com/books/edition/Introduction_to_Probability_with_Texas_H/BZW_DQAAQBAJ?hl=en&gbpv=0), 
an R [package](https://cran.r-project.org/package=holdem), and a
semester-long [class](http://www.stat.ucla.edu/~frederic/100A/sum20/) at UCLA on
learning probability using poker examples. Poker is the only example I can
think of that utilizes all three conceptions of probability:

1. Classical: The deck is discrete and finite and each card (remaining) in a shuffled
  deck has the same probability of being revealed next
2. Bayesian: Within a particular hand, the players essentially have a decision
  theory problem where the conditional probability of winning the hand
  given the visible cards, the betting, body language, past hands, etc. is governed 
  by Bayes' Rule and is clearly subjective
3. Frequentist: To evaluate a poker strategy or poker player, you really need to
  consider the results over thousands of hands from shuffled deck

But what does poker have to do with the QMSS? Poker is an example of an 
incomplete information, non-cooperative, zero-sum game that 
[motivated](https://timharford.com/2006/05/the-poker-machine/) 
John von Neumann to develop game theory. In is interesting
that social scientists often assume that _agents_ in a formal model act
as if they were Bayesian but rarely do those social scientists _estimate_ the
parameters of those models using Bayesian methods. In addition, 
Maria Konnikova, who has a Ph.D. in psychology from Columbia, has
written a [book](https://www-nature-com.ezproxy.cul.columbia.edu/articles/d41586-020-01840-5)
on how learning to be a professional poker player affected her and
how psychological limitations of individuals prevent them from playing poker
optimally and from optimally doing other things in life.

As professional social scientists, we want to look at social science
the way professional poker players look at poker. In both cases,
there are known and unknown quantities as well as models relating
one to the other. We can use Bayes' Rule to obtain the conditional
probability of what we don't know given what we do know and then
use the probabilities to make the decision or take the action that has the
highest expected utility. Frequentist and supervised learning approaches
do not yield the probability of what we don't know given what they do know
and thus do not provide a sound basis for making decisions or taking
actions as a result of research.

## Pot Odds

Mike Postle needs to have slightly more than a one-in-three chance
(specifically $\frac{125}{125 + 235} \approx 0.347$) to win the
pot given that Marle Cordeiro has an above median hand. If Marle Cordeiro
had a hand like a pair of Aces, Kings, or Queens, then the probability
that Mike Postle would win with \Qd\Jh \ is much smaller than one-in-three
(actually smaller than one-in-five) but that is only 
$3 \times {4 \choose 2} = 18$ of the $\frac{{52 \choose 2}}{2} = 663$ above-median
combinations that Marle Cordeiro could have. In addition, if Marle Cordeiro has a
Queen with either an Ace or a King, that is not good for Mike Postle but that
is only $2 \times 4^2 = 32$ combinations. Against the bulk of the above-median
combinations that Marle Cordeiro could have, a \Qd\Jh \ will have a $0.4$ to $0.5$
probability of winning so the marginal probability is well over $0.347$.

## Probability of a Straight

The flop must contain a Jack and two other cards that are adjacent to her
\Qs\tend \ in order for Marle Cordeiro to make a straight. Since there are
$4$ Jacks among the $50$ unobserved cards, the probability of a Jack is simply
$\frac{4}{50} = 0.08$. Then, for example, the probability of getting a nine
and an eight on the flop, given that a Jack is already turned over, is
$\frac{4}{49} \times \frac{4}{48} \approx 0.0068$. But the probability of
an Ace and a King, given that a Jack is already turned over is the same, 
as is the probability of a King and a nine. Also, there are $3! = 6$ orders
in which the three cards that complete Marle Cordeiro's straight on the
flop can appear, so from Marle Cordeiro's perspective, the probability of
making a straight on the flop given that she has \Qs\tend \ as hole cards is
$$\frac{4}{50} \times 3 \frac{4}{49}\frac{4}{48} \times 3! \approx 0.0098$$

## First Mover Disadvantage

The first player's action can be used by the last player to update their
beliefs about the first player's hole cards. Conversely, the first player
cannot update their beliefs about the last player's hole cards at the time
that the first player has to make a decision.

As a result, in most cases where the small blind or the big blind is the first to act, 
the Game Theory Optimal strategy is for them to check with _all_ combinations of hole 
cards (that had not previously been folded) in order to prevent subsequent players
from updating their beliefs.

## The Turn

If you answered the part about the probability of Marle Cordeiro is _dealt_ a
combination of cards unconditionally or conditional on the four cards that were
turned face-up in the middle, that is acceptable as long as your math corresponds to
it. But I intended it to be answered before the flop but conditional on Mike Postle's
hole cards.

From Mike Postle's perspective, the probability that Marle Cordeiro is dealt a 4
with a 9, 8, or Jack from a deck that is already lacking the \Qd \ and the \Jh \ is
$\frac{4}{50} \times 3 \frac{4}{49} \times 2 \approx 0.04$. But if Marle Cordeiro had a 4 
with a 9, 8, or Jack, that would be considerably worse than the median combination of 
a Queen and a 8. Thus, the conditional probability that Marle Cordeiro raises to \$150 
at her first opportunity given that she has a 4 with a 9, 8, or Jack is essentially zero.
  
From Mike Postle's perspective, the probability that Marle Cordeiro is dealt a pair of
4s from a deck that is already lacking the \Qd \ and the \Jh \ is
$\frac{4}{50} \times \frac{3}{49} \times 2 \approx 0.01$. And if Marle Cordeiro had a
pair of 4s (or any other pair) that would be better than the median combination of a
Queen and a 8, mostly due to the chance that if you then make three or four of a kind, 
then you are likely to win a lot of chips. Thus, the conditional probability that Marle
Cordeiro raises to \$150 at her first opportunity given that she has a pair of 4s is
essentially one. According to Bayes' Rule, the posterior probability that Marle Cordeiro 
has a pair of 4s given that she raised to \$150 at her first opportunity is
$$\frac{\frac{4}{50} \times \frac{3}{49} \times 2 \times 1}{1/2} \approx 0.02$$
where we have used the probability of being dealt a pair of 4s as the prior, a 
conditional probability of raising of $1$, and a marginal probability of raising of
$\frac{1}{2}$. In other words, the posterior probability that Marle Cordeiro has
a pair of 4s is double the prior probability.
  
However, small pairs are likely to be among the 20\% of combinations of hole cards
(that had not already been folded) where Marle Cordeiro would check after Mike Postle
checks when the flop comes down as \nined\eigs\Jd, since a pair of 4s is probably
trailing a bare majority of the hands that Mike Postle would not have already folded.
If there is less than a 4-in-10 chance that Marle Cordeiro would raise \$200 on the
turn given that she has a pair of 4s and a 8-in-10 chance that Marle Cordeiro would 
raise \$200 on the turn marginally, then the conditional probability that she has
a pair of 4s would be back below the 1-in-100 chance that she got dealt a pair of 4s.

## Mike Postle's Fold

Veronica Brill is puzzled, to put it mildly, why Mike Postle would fold after
Marle Cordeiro bet \$200 on the turn. If the pot odds of $\frac{235}{125} = 1.88$ were 
such that Mike Postle should call before the flop and the pot odds of $\frac{360}{200} = 1.8$
where such that Mike Postle should call after the flop (both of which are justified),
then why should Mike Postle _not_ call after the turn with a pot odds of 
$\frac{760}{600} \approx 1.27$ when there is very little chance that the \fours \ vaulted 
Marle Cordeiro into the lead with three of a kind against a pair of Jacks?

Mike Postle would only be trailing Marle Cordeiro if she had

* A pair of Aces (6 combinations), a pair of Kings (6 combinations), or a 
  pair of Queens (3 combinations, since he has \Qd), which are consistent with her
  betting and would give her a higher pair
* A pair of 9s (3 combinations), a pair of 8s (3 combinations),
  a pair of Jacks (1 combination, since he has \Jh), which are consistent with her
  betting and would give her a three of a kind
* A pair of 4s (3 combinations) would also give her three of a kind but is not that 
  all that consistent with her betting, and neither is any two pair except perhaps
  a nine and an eight of the same suit (2 combinations since a nine and a eight
  of different suits have been turned over)
* A Jack with an Ace or a King (8 combinations, since he has \Jh), which are 
  consistent with her betting and would give her the same pair but a higher kicker
* A straight (12 combinations, since he has the \Qh \ and she would have folded with 
  any 10 with a 7)

In other words, although Mike Postle could have been trailing all along, he has
every reason to _expect_ that he would win a lot of chips if he calls versus none
if he folds. Moreover, if Mike Postle regularly makes decisions that are as bad as
this one, then he is very unlikely to make any money playing poker, much less a
lot of money almost every time that he has played (on Stones Live) over the last several 
months. Mike Postle's fold is rational only if he somehow has additional information
--- beyond the cards in the middle and the betting history --- that Marle Cordeiro
happens to have one of the few combinations of hole cards that is better than his
\Qd\Jh \. Perhaps that additional information is a tell, but since it seems implausible
that Mike Postle would be able to pick up on a tell of a professional poker in 
such a short amount of time, the prospect that Mike Postle's additional information
is illegal becomes more concerning.

## Statistical Considerations

This closely resembles the logic of Frequentist testing of a null hypothesis. In this
case we have an empirical distribution of blue dots in a two-dimensional space where
the players are presumably not cheating (if for no other reason that it is difficult
to pull off) and are mostly independent of each other. These players tend to win or
lose at most 100 big blinds per hundred hands. The question becomes should we reject
the "null" hypothesis that the red and / or orange dots were generated by players
that were playing fairly. In the case of the red dot, "potripper" managed to win
at a much higher rate than a typical player while folding less than 1-in-10 hands.
In other words, he was betting with much _worse_ hole cards than what most players
bet with and was somehow winning five times more than the best players. Thus, it
when he admitted to cheating, it was not much of a surprise. According to the
plot, Mike Postle was somehow winning ten times more than the best players while
folding at his first opportunity about half as often. Thus, almost all poker
players that have seen this graph would reject a "null" hypothesis that Mike Postle
was playing fairly, and it would make sense that Mike Postle would claim that the
graph overstates how much he won since July 2018.

## The Phone

In contrast, the evidence from Mike Postle's phone is a good example of Bayesian
updating. In general, we could write Bayes' Rule as
$$\Pr\left(\mbox{cheating} \mid \mbox{evidence}\right) = 
\frac{\Pr\left(\mbox{cheating}\right) \times \Pr\left(\mbox{evidence} \mid \mbox{cheating}\right)}
{\Pr\left(\mbox{cheating}\right) \times \Pr\left(\mbox{evidence} \mid \mbox{cheating}\right) + 
 \Pr\left(!\mbox{cheating}\right) \times \Pr\left(\mbox{evidence} \mid !\mbox{cheating}\right)}$$
where the exclamation point reads as "not". Again, the prior probability that 
Mike Postle (or anyone else) is cheating, $\Pr\left(\mbox{cheating}\right)$ is 
presumably low if for no other reason that it is difficult to pull off.

However, if Mike Postle were cheating by somehow seeing opponents' hole cards on
his phone, then it would make sense that he would put his phone in a location
where he could see it but no one else (and no cameras) could see it. Thus, the
second term in the numerator of Bayes' Rule is close to one. Conversely, if Mike
Postle were not cheating, then it seems unlikely that he would start putting
his phone on his chair rather than on the table. Indeed, it seems that all 
non-sketchy uses of a phone would be more difficult when his phone is in the chair.
Thus, the second term in the denominator of Bayes' Rule is not very big and 
the prior probability that Mike Postle is updated to something much closer
to one as a result of this evidence. 

Indeed, we could define a "smoking gun" as evidence where one term in the denominator 
of Bayes Rule is close to zero and the other is close to one, in which case the posterior 
probability would be close to one for any non-dogmatic prior. Placing the
phone on the chair, by itself, is perhaps not quite a smoking gun since Mike Postle
has claimed that he was looking at private pictures that were sent to him,
which is one of the few other reasons he would obscure his phone. Conversely,
Stones claimed to have conducted a thorough investigation and found no evidence
that Mike Postle cheated, which did not result in poker players updating their
beliefs very much. If Mike Postle were cheating, it would make sense that 
Stones would proclaim that it had not found such evidence, because if there
were evidence it would have presumably implicated a Stones' employee in which
case Stones would be liable and quite possibly shut down by the state of California.

Similarly, if Mike Postle were cheating by intercepting the RFID information
produced by the table, then it would make sense for him to have some software
on his phone that was similar to the software the production room at Stones
uses to read the cards and merge their images into the video stream. Conversely,
if if Mike Postle were not cheating, why would he be looking at a blue but
otherwise blank screen on his phone? He is not getting status updates on Facebook
or ordering an Uber or using any other recognizable app. Indeed, why would
there be an app that shows nothing but a blue screen and if there were such an
app, why would anyone download or use it? 

Finally, if Mike Postle were cheating by intercepting the RFID information
produced by the table, then he would know whether the table had malfunctioned
on a hand where it was not reading all of the hole cards. Conversely, if Mike
Postle were not intercepting the RFID information, then he would not know that
the table had malfunctioned on a hand. Thus, if indeed he asked a Stones' 
employee about a hand where the table malfunction by not reading all the hole
cards, that would be close to a smoking gun.

## The Court Cases

This seems like a thorny issue, albeit one that recently came up again when
Dominion, which makes voting machines, sued Rudy Giuliani for claiming that
their voting machines were rigged against Donald Trump. On one hand, it would not 
make sense to punish people for saying things they believed to be true (even if they 
were false), and Bayesians feel that beliefs should be updated according to Bayes' Rule.
However, any Bayesian defendant could claim something like "My prior was
such that the plaintiff was defrauding people, and thus it was rational
for me to believe so, even though the evidence was not strong." In addition,
that prior could be based on prejudice or past experience being defrauded
by people _other_ than the plaintiff, which the legal system would not want
to be a factor in deciding the case at hand.

Similarly, Bayesian defendants might not be that sure what they are saying
is true, but their utility functions could be such that it is rational for
them to act as if it were true. Indeed, Veronica Brill's original thread
about Mike Postle included this [tweet](https://twitter.com/Angry_Polak/status/1178014856793845762):

> Am I sure that this player is cheating? No. Do I think that there is a greater 
than zero \% chance that he is? Yes.

If Veronica Brill's utility function is such that she does not want
herself or other people to continue to lose money to someone who is not
playing fairly, then it would make sense to raise these concerns about
Mike Postle for some sufficiently high posterior probability that he is
cheating. But the utility function, like the prior distribution, is personal
and the legal system is ostensibly impersonal.

Even in criminal cases more broadly, having Bayesian jurors would raise
issues. Judges do not want jurors to condition on _all_ the information
they might have; only the evidence that is (admissibly) presented during
the trial. Prosecutors tend to win at least 90\% of cases that go to 
trial, so (assuming no prosecutorial misconduct) a Bayesian juror might
well have a prior that the defendant is guilty with probability $0.9$
before any evidence is presented, although a judge would dismiss a 
juror that admitted to having such a prior.

All that said, the Frequentist perspective on probability does not
seem adequate for legal proceedings either. It is irrelevant if the
legal system only convicts 5\% of innocent defendants (or has
some other known Type I error rate). Jurors have to decide whether 
_this_ particular defendant is guilty when nothing is randomized,
and all of the uncertainty is due to incomplete information rather
than a finite $N$.
