---
title: "GR5065 Homework 1"
date: "Due January 26, 2021 at 8PM New York Time"
author: "Your Name"
output: 
  pdf_document: 
    latex_engine: xelatex
    number_sections: yes
---

```{r, setup}
# call the set.seed function once here to make the knitting conditionally deterministic
```

# Bowling

This is an R chunk that will be executed so that `F` is available to call in subsequent R chunks
and the output will be included in your knitted PDF document:
```{r}
# computes the x-th Fibonacci number without recursion and with vectorization
F <- function(x) {
  stopifnot(is.numeric(x), all(x == as.integer(x)))
  sqrt_5 <- sqrt(5) # defined once, used twice
  golden_ratio <- (1 + sqrt_5) / 2
  return(round(golden_ratio ^ (x + 1) / sqrt_5))
}
```

## R Implementation

Make another R chunk here where you define the `Pr` function

## Admissibility

## Simulating a Game of Bowing

### First Roll

### Second Roll

### Probability of a Frame

### Game

### Probability of a Game

# Poker

## Pot Odds

## Probability of a Straight

## First Mover Disadvantage

## The Turn

## Mike Postle's Fold

## Statistical Considerations

## The Phone

## The Court Cases
